Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-10-03 paper/paper-infer.tex
typo

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesR where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
In that paper, they found consistent results with BayesR, although the original BayesR publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}. In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).

>>>>

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesS where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
In that paper, they found consistent results with BayesS, although the original BayesS publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}. In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).
